Physcraper: a python package for continual update of evolutionary estimates using the Open Tree of Life

### 2020-08-18 2-body.Rmd
results fig

An online literature review in June 2020 (google scholar search for "ilex phylogeny") reveals that there are several published phylogenetic trees showing relationships within the hollies [@cuenoud2000molecular; @manen2010history; @setoguchi2000intersectional; @selbach2009new], but only two have their data available publicly [@gottlieb2005molecular; @yao2020phylogeny].
@gottlieb2005molecular made original tree and alignment data available in TreeBASE. The "Gottlieb2005" tree sampling 48 species was added to the [OpenTree Phylesystem](https://tree.opentreeoflife.org/curator/study/edit/pg_2827/?tab=home) and its information has been integrated into [OpenTree's synthetic tree](https://devtree.opentreeoflife.org/opentree/opentree12.3@mrcaott68451ott89474/Ilex-theizans--Ilex-dumosa).

The most recent *Ilex* tree from @yao2020phylogeny, has been made available in the [OpenTree Phylesystem](https://tree.opentreeoflife.org/curator/study/view/ot_1984) and in the [DRYAD repository](https://datadryad.org/stash/dataset/doi:10.5061/dryad.k0p2ngf4x). The "Yao2020" tree, is the best sampled phylogenetic tree yet available for the hollies, with 175 tips.
<!-- However, it has not been added to OpenTree's synthtetic tree yet. -->
This makes it a great case to highlight the functionalities of Physcraper and assess its limitations.
<!-- we know that the sequences of the Yao2020 tree have been made available on the GenBank database. Hence, we expect that updating the Gottlieb2005 tree should at least recover the same species sampled in the Yao2020 tree. -->
A tutorial as well as illustrated examples of functions implemented on each step of the analysis are available in Physcraper's documentation website.
Figure \@ref(fig:results) shows results from updating one alignment from the Gottlieb2005 tree. It contains all 41 distinct taxa from the original Gottlieb2005 tree plus 231 new tips, contributing phylogenetic data to 84 additional taxa within *Ilex*. The tree is 99% resolved, however, a large portion of branches are negligibly small (133 branches < 0.00001 substitution rate units), appearing as unresolved nodes in the figure.

@@

```{r results, echo=FALSE, fig.cap="A phylogenetic tree is the main result of a Physcraper analysis. The tree results from updating an alignment that was used to generate a tree of 48 tips belonging to the genus Ilex published in Gottlieb et al. 2005 (black tips). Physcraper added 231 new sequences (red tips) corresponding to 84 distinct taxa that were not sampled in the original alignment. Physcraper obtained sequences from the GenBank database via BLAST, filtered them following specified criteria, aligned them to the original alignment using MUSCLE and performed a phylogenetic reconstruction using RAxML as well as a conflict analysis.", out.width = '100%', fig.retina= 2}

knitr::include_graphics("docs/figs/ilex-plot-tips-and-branches-1.png")

@@

It has THIS many resolved nodes, and THIS MANY nodes with bootsrap values above 95%

The original tree has THIS MANY resolved nodes

The alignments have THIS MANY columns and THIS AMOUNT of missing data.

DESCRIBE RESULTS: SUMMARY OF NEW TAXA FOUND RELATIVE TO ORIGINAL TREE AND RELATIVE TO OpenTree
RF DISTANCE INTERPRETATION
HOW MUCH TIME THE BLAST RUN TOOK
ML ESTIMATES OF UPDATED TREE VS ORIGINAL TREE

FIGURE: FACE TO FACE ORIGINAL VS UPDATED PHYLOGENY, IN RED NEW TAXA NOT IN OpenTree.

@@

>>>>

An online literature review in June 2020 (google scholar search for "ilex phylogeny") reveals that there are several published phylogenetic trees showing relationships within the hollies [@cuenoud2000molecular; @manen2010history; @setoguchi2000intersectional; @selbach2009new], but only two have their data available publicly [@gottlieb2005molecular; @yao2020phylogeny].
@gottlieb2005molecular made original tree and alignment data available in [TreeBASE](https://treebase.org/treebase-web/search/study/summary.html?id=1091). The "Gottlieb2005" tree sampling 41 species was added to the [OpenTree Phylesystem](https://tree.opentreeoflife.org/curator/study/edit/pg_2827/?tab=home) and its information has been integrated into [OpenTree's synthetic tree](https://devtree.opentreeoflife.org/opentree/opentree12.3@mrcaott68451ott89474/Ilex-theizans--Ilex-dumosa).

The most recent *Ilex* tree from @yao2020phylogeny, has been made available in the [OpenTree Phylesystem](https://tree.opentreeoflife.org/curator/study/view/ot_1984) and in the [DRYAD repository](https://datadryad.org/stash/dataset/doi:10.5061/dryad.k0p2ngf4x). The "Yao2020" tree, is the best sampled phylogenetic tree yet available for the hollies, with 175 tips.
<!-- However, it has not been added to OpenTree's synthtetic tree yet. -->
This makes it a great case to highlight the functionalities of Physcraper and assess its limitations.
<!-- we know that the sequences of the Yao2020 tree have been made available on the GenBank database. Hence, we expect that updating the Gottlieb2005 tree should at least recover the same species sampled in the Yao2020 tree. -->
A tutorial as well as illustrated examples of functions implemented on each step of the analysis are available in Physcraper's documentation website.
Figure \@ref(fig:results) shows results from a Physcraper analysis applied to an alignment of internal transcribed spacer DNA regions (ITS) from @gottlieb2005molecular. Physcraper ran on a local BLAST database, on a laptop Linux computer for 19hrs 45min to finish BLAST and RAxML analyses. Bootstrap analyses took an additional 13hrs.
The updated Gottlieb2005 tree contains all 41 distinct taxa from the original study plus 231 new tips, contributing phylogenetic data to 84 additional taxa within *Ilex*. The tree is 99% resolved, however, a large portion of branches are negligibly small (133 branches < 0.00001 substitution rate units), appearing as unresolved nodes (Fig. \@ref(fig:results)).
<!-- ADD percentage of nodes with bootsrap values above 95%  and below 50%?-->
For comparison, the Yao2020 tree also contains all 41 distinct taxa from the original Gottlieb2005 study,
<!-- has less new tips (134), but  -->
and contributes phylogenetic data to 134 additional taxa within *Ilex*, from which
`r 135-68` are also in updated Gottlieb2005.
<!-- The alignments have THIS MANY columns and THIS AMOUNT of missing data. -->
<!-- RF DISTANCE INTERPRETATION -->
<!-- ADD ML ESTIMATES OF UPDATED TREE VS ORIGINAL TREE -->

@@

```{r results, echo=FALSE, fig.cap="A phylogenetic tree is the main result of a Physcraper analysis.", out.width = '100%', fig.retina= 2}

# knitr::include_graphics("docs/figs/ilex-plot-tips-and-branches-1.png")
knitr::include_graphics("docs/figs/ilex-results.png")

@@

@@

Figure \@ref(fig:results) caption continued: The tree results from updating an alignment that was used to generate a tree of 48 tips of the genus Ilex, published in Gottlieb et al. 2005 (black tips). Physcraper added 231 new sequences (red tips) corresponding to 84 distinct taxa that were not sampled in the original alignment. Physcraper obtained sequences from the GenBank database via BLAST, filtered them following specified criteria, aligned them to the original alignment using MUSCLE and performed a phylogenetic reconstruction using RAxML, as well as a conflict analysis using the OpenTree tools.

### 2020-08-18 docs/submission-mee-2020-08-17.tex
results fig

\textbf{Word count}: 3660

@@

To illustrate the utility of Physcraper we will use a scenario in which a user is interested in a particular group, e.g., the genus \emph{Ilex}, which encompasses between 400-700 living species, and is the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering plants, commonly known as ``hollies''.

An online literature review in June 2020 (google scholar search for ``ilex phylogeny'') reveals that there are several published phylogenetic trees showing relationships within the hollies (Cuénoud \emph{et al.} \protect\hyperlink{ref-cuenoud2000molecular}{2000}; Setoguchi \& Watanabe \protect\hyperlink{ref-setoguchi2000intersectional}{2000}; Selbach-Schnadelbach \emph{et al.} \protect\hyperlink{ref-selbach2009new}{2009}; Manen \emph{et al.} \protect\hyperlink{ref-manen2010history}{2010}), but only two have their data available publicly (Gottlieb \emph{et al.} \protect\hyperlink{ref-gottlieb2005molecular}{2005}; Yao \emph{et al.} \protect\hyperlink{ref-yao2020phylogeny}{2020}).
Gottlieb \emph{et al.} (\protect\hyperlink{ref-gottlieb2005molecular}{2005}) made original tree and alignment data available in TreeBASE. The ``Gottlieb2005'' tree sampling 48 species was added to the \href{https://tree.opentreeoflife.org/curator/study/edit/pg_2827/?tab=home}{OpenTree Phylesystem} and its information has been integrated into \href{https://devtree.opentreeoflife.org/opentree/opentree12.3@mrcaott68451ott89474/Ilex-theizans--Ilex-dumosa}{OpenTree's synthetic tree}.
The most recent \emph{Ilex} tree from Yao \emph{et al.} (\protect\hyperlink{ref-yao2020phylogeny}{2020}), has been made available in the \href{https://tree.opentreeoflife.org/curator/study/view/ot_1984}{OpenTree Phylesystem} and in the \href{https://datadryad.org/stash/dataset/doi:10.5061/dryad.k0p2ngf4x}{DRYAD repository}. The ``Yao2020'' tree, is the best sampled phylogenetic tree yet available for the hollies, with 175 tips.
This makes it a great case to highlight the basic functionalities of Physcraper: we know that the sequences of the Yao2020 tree have been made available on the GenBank database. Hence, we expect that updating the Gottlieb2005 tree should at least recover the same species sampled in the Yao2020 tree.
A tutorial as well as illustrated examples of functions implementated on each step of the Physcraper framework are available in the documentation website.

\begin{figure}
\includegraphics[width=1\linewidth]{docs/figs/ilex-plot-tips-and-branches-1} \caption{A phylogenetic tree is the main result of a Physcraper analysis. In this example, Physcraper updated an alignment of 48 samples of the genus Ilex from Gottlieb et al. 2005 (black tips) and added 222 new samples from the GenBank database (red tips), corresponding to 78 new taxa, aligned the new seqeunces to the original alignment using MUSCLE and performed the phylogenetic reconstruction using RAxML.}\label{fig:ilex-updated-tree}
\end{figure}

The Physcraper updated tree has 57 original tips, representing 51 unique OTT taxa, and 222 new tips representing 78 new OTT taxa added to the original tree.

It has THIS many resolved nodes, and THIS MANY nodes with bootsrap values above 95\%

The original tree has THIS MANY resolved nodes

The alignments have THIS MANY columns and THIS AMOUNT of missing data.

DESCRIBE RESULTS: SUMMARY OF NEW TAXA FOUND RELATIVE TO ORIGINAL TREE AND RELATIVE TO OpenTree
RF DISTANCE INTERPRETATION
HOW MUCH TIME THE BLAST RUN TOOK
ML ESTIMATES OF UPDATED TREE VS ORIGINAL TREE

FIGURE: FACE TO FACE ORIGINAL VS UPDATED PHYLOGENY, IN RED NEW TAXA NOT IN OpenTree.

@@

@@

COMPARE WITH PERFORMANCE OF OTHER PIPELINES FOR SEQUENCE SCRAPING
WHY WE DID NOT MAKE A BENCHMARK COMPARISON

@@

We thank members of the ``short bar'' Science and Engineering Building 1, University of California Merced, joint lab paper discussion meeting for valuable comments on a first version of this manuscript.

>>>>

\textbf{Word count}: 3745

@@

To illustrate the utility of Physcraper we propose a scenario in which a user is interested in phylogenetic relationships within the genus \emph{Ilex}. Commonly known as ``hollies'', the genus encompasses between 400-700 living species, and is the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering plants.

An online literature review in June 2020 (google scholar search for ``ilex phylogeny'') reveals that there are several published phylogenetic trees showing relationships within the hollies (Cuénoud \emph{et al.} \protect\hyperlink{ref-cuenoud2000molecular}{2000}; Setoguchi \& Watanabe \protect\hyperlink{ref-setoguchi2000intersectional}{2000}; Selbach-Schnadelbach \emph{et al.} \protect\hyperlink{ref-selbach2009new}{2009}; Manen \emph{et al.} \protect\hyperlink{ref-manen2010history}{2010}), but only two have their data available publicly (Gottlieb \emph{et al.} \protect\hyperlink{ref-gottlieb2005molecular}{2005}; Yao \emph{et al.} \protect\hyperlink{ref-yao2020phylogeny}{2020}).
Gottlieb \emph{et al.} (\protect\hyperlink{ref-gottlieb2005molecular}{2005}) made original tree and alignment data available in \href{https://treebase.org/treebase-web/search/study/summary.html?id=1091}{TreeBASE}. The ``Gottlieb2005'' tree sampling 41 species was added to the \href{https://tree.opentreeoflife.org/curator/study/edit/pg_2827/?tab=home}{OpenTree Phylesystem} and its information has been integrated into \href{https://devtree.opentreeoflife.org/opentree/opentree12.3@mrcaott68451ott89474/Ilex-theizans--Ilex-dumosa}{OpenTree's synthetic tree}.

The most recent \emph{Ilex} tree from Yao \emph{et al.} (\protect\hyperlink{ref-yao2020phylogeny}{2020}), has been made available in the \href{https://tree.opentreeoflife.org/curator/study/view/ot_1984}{OpenTree Phylesystem} and in the \href{https://datadryad.org/stash/dataset/doi:10.5061/dryad.k0p2ngf4x}{DRYAD repository}. The ``Yao2020'' tree, is the best sampled phylogenetic tree yet available for the hollies, with 175 tips.
This makes it a great case to highlight the functionalities of Physcraper and assess its limitations.
A tutorial as well as illustrated examples of functions implemented on each step of the analysis are available in Physcraper's documentation website.
Figure \ref{fig:results} shows results from a Physcraper analysis applied to an alignment of internal transcribed spacer DNA regions (ITS) from Gottlieb \emph{et al.} (\protect\hyperlink{ref-gottlieb2005molecular}{2005}). Physcraper ran on a local BLAST database, on a laptop Linux computer for 19hrs 45min to finish BLAST and RAxML analyses. Bootstrap analyses took an additional 13hrs.
The updated Gottlieb2005 tree contains all 41 distinct taxa from the original study plus 231 new tips, contributing phylogenetic data to 84 additional taxa within \emph{Ilex}. The tree is 99\% resolved, however, a large portion of branches are negligibly small (133 branches \textless{} 0.00001 substitution rate units), appearing as unresolved nodes (Fig. \ref{fig:results}).
For comparison, the Yao2020 tree also contains all 41 distinct taxa from the original Gottlieb2005 study,
and contributes phylogenetic data to 134 additional taxa within \emph{Ilex}, from which
67 are also in updated Gottlieb2005.

\begin{figure}
\includegraphics[width=1\linewidth]{docs/figs/ilex-results} \caption{A phylogenetic tree is the main result of a Physcraper analysis.}\label{fig:results}
\end{figure}

@@

Figure \ref{fig:results} caption continued: The tree results from updating an alignment that was used to generate a tree of 48 tips of the genus Ilex, published in Gottlieb et al.~2005 (black tips). Physcraper added 231 new sequences (red tips) corresponding to 84 distinct taxa that were not sampled in the original alignment. Physcraper obtained sequences from the GenBank database via BLAST, filtered them following specified criteria, aligned them to the original alignment using MUSCLE and performed a phylogenetic reconstruction using RAxML, as well as a conflict analysis using the OpenTree tools.

@@

\emph{COMPARE WITH PERFORMANCE OF OTHER PIPELINES FOR SEQUENCE SCRAPING}
\emph{WHY WE DID NOT MAKE A BENCHMARK COMPARISON?}

@@

We thank members of the ``short bar'' Science and Engineering Building 1, UCM, joint lab paper discussion group for valuable comments on a first version of this manuscript.
